---
title: "Vaccination rate mini project"
author: "Pham Vo"
date: "3/6/2022"
output:
  pdf_document: 
    latex_engine: xelatex
---

## Getting Started

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```
> Q1. What column details the total number of people fully vaccinated? persons_fully_vaccinated

> Q2. What column details the Zip code tabulation area? zip_code_tabulation_area

> Q3. What is the earliest date in this dataset? 2021-01-05

> Q4. What is the latest date in this dataset? 2022-03-01	 

```{r}
# As we have done previously, let’s call the skim() function from the skimr package to get a quick overview of this dataset

library(skimr)
skimr::skim(vax)
```


> Q5. How many numeric columns are in this dataset? 10

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

```{r}
round(sum(is.na(vax$persons_fully_vaccinated))/length(vax$persons_fully_vaccinated)*100,2)
```

## Working with dates

```{r}
library(lubridate)
today()
```

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
# Now we can do math with dates. For example: How many days have passed since the first vaccination reported in this dataset

today() - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset?

```{r}
# Using the last and the first date value we can now determine how many days the dataset span

vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
unique.dates <- (unique(vax$as_of_date))
num.unique.dates <- length(unique.dates)
num.unique.dates
```

## Working with ZIP codes

```{r}
library(zipcodeR)

# find the centroid of the La Jolla 92037 (i.e. UC San Diego) ZIP code area
geocode_zip('92037')
```
```{r}
# Calculate the distance between the centroids of any two ZIP codes in miles
zip_distance('92037','92109')
```
```{r}
# pull census data about ZIP code areas (including median household income etc.)
reverse_zipcode(c('92037', "92109") )
```
```{r}
# Pull data for all ZIP codes in the dataset
#zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

## Focus on the San Diego area

```{r}
# Subset to San Diego county only areas
sd <- vax[vax$county == "San Diego", ]
```

```{r}
# Using dplyr the code would look like this

library(dplyr)
sd <- filter(vax, county == "San Diego")
nrow(sd)
```

```{r}
# subsetting across multiple criteria - for example all San Diego county areas with a population of over 10,000

sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(vax$zip_code_tabulation_area))
```

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
largest_plus12_population <- which.max(vax$age12_plus_population)
vax$zip_code_tabulation_area[largest_plus12_population]
```

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-22”?

```{r}
sd_20220222 <- filter(vax, county == "San Diego" & as_of_date == "2022-02-22")
round(mean(na.omit(sd_20220222$percent_of_population_fully_vaccinated))*100,2)
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-02-22”?

```{r}
pfv_20220222 <- filter(vax, as_of_date == "2022-02-22")
library(ggplot2)
ggplot(pfv_20220222, aes(percent_of_population_fully_vaccinated)) +
  geom_histogram() +
  labs(x="Percent of Population Fully Vaccinated in a Zip Code Area", y="Count (Zip code areas)", title = "Histogram of Vaccination Rates Across San Diego County", subtitle = "As of 2022-02-22")
```

## Focus on UCSD/La Jolla

```{r}
#UC San Diego resides in the 92037 ZIP code area and is listed with an age 5+ population size of 36,144

ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(as_of_date, percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", title = "Vaccination rate for La Jolla CA 92109")
```

## Comparing to similar sized areas

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-02-22")

#head(vax.36)
```
> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
vax.mean.36 <- mean(na.omit(vax.36$percent_of_population_fully_vaccinated))

ggplot(ucsd) +
  aes(as_of_date, percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", title = "Vaccination rate for La Jolla CA 92109") +
  geom_hline(aes(yintercept=vax.mean.36), linetype = "dashed", color = "red")
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”?

```{r}
summary(vax.36)
```

> Q18. Using ggplot generate a histogram of this data

```{r}
ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() +
  xlim(c(0,1)) +
  labs(x="Percent Vaccinated", y="Count")
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
zip_92040 <- vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)

zip_92040 < vax.mean.36
```

```{r}
zip_92109 <- vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)

zip_92109 < vax.mean.36
```

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1)) +
  labs(x= "Date", y= "Percent Vaccinated",
       title= "Vaccination rate across California",
       subtitle= "Only areas with a population above 36k are shown") +
  geom_hline(yintercept = vax.mean.36, linetype= "dashed")
```

```{r}
sessionInfo()
```

