---
title: "DESeq2 mini project"
author: "Pham Vo"
date: "2/27/2022"
output:
  pdf_document: default
  html_document: default
---

```{r}
library(DESeq2)
```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
head(colData)
```
```{r}
# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```

```{r}
# Note we need to remove the odd first $length col
countData <- as.matrix(countData[,-1])
head(countData)
```
```{r}
# Filter count data where you have 0 read count across all samples.
countData = countData[rowSums(countData[])>0,]
head(countData)
```
## Running DESeq2

```{r}
library(DESeq2)
citation("DESeq2")
```

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```
```{r}
dds
```
```{r}
res = results(dds, alpha=0.1)
summary(res)
```

## Volcono plot

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj<0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

## Adding gene annotation

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```

```{r}
# Reorder these results by adjusted p-value and save them to a CSV file in your current project directory

res = res[order(res$pvalue),]
write.csv(res[order(res$padj),], "deseq_results_1.csv")
```


## Pathway Analysis

```{r}
library(pathview)
```
```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```


```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```


```{r}
# Look at the first few down (less) pathways
head(keggres$less)
```
```{r}
head(keggres$greater)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```
```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids

pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

## Reactome Analysis
> Reactome is database consisting of biological molecules and their relation to pathways and processes

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

> Then, to perform pathway analysis online go to the Reactome website (https://reactome.org/PathwayBrowser/#TOOL=AT). Select “choose file” to upload your significant gene list. Then, select the parameters “Project to Humans”, then click “Analyze”.


## GO online

> To perform Gene Set GO Enrichment online go to the website http://www.geneontology.org/page/go-enrichment-analysis. Paste your significant gene list from section 4. Then, select "biological process" and "homo sapiens", and click submit.

```{r}
sessionInfo()
```

