---
title: "Machine Learning"
author: "Pham Vo"
date: "2/13/2022"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Clustering with kmeans and hclust()

We will begin by making up some data to cluster.
shortcut to insert code chunk: option+command+i

```{r}
tmp <- c(rnorm(30,3), rnorm(30,-3))
x <- cbind(x=tmp,y=rev(tmp))
x
```

## Run kmeans()
```{r}
k <- kmeans(x, centers=2, nstart=20)
k
```
> Q. what size is each cluster?

```{r}
k$size
```

> Q. Cluster centers

```{r}
k$centers
```

> Q. Membership vector

```{r}
k$cluster
```

> Plot our data with the clustering result 

```{r}
plot(x, col=k$cluster)
points(k$centers, col="blue", pch=16, cex=1.5)
points(x[42,1], x[42,2], col="darkgreen", pch = 16)
points(x[19,1], x[19,2], col="brown", pch = 16)
```

## hclust()

```{r}
hc <- hclust(dist(x))
hc
```
There is a cool and useful plot method for hclust()

```{r}
plot(hc)
```

# Principal Component Analysis
Data import

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
x
```
#Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
ncol(x)
nrow(x)
```


Let's make some typical plots

```{r}
mycols <- rainbow(nrow(x))
barplot(as.matrix(x), col=mycols)
```

Make it side by side i.e. not stacked

```{r}
barplot(as.matrix(x), col=mycols, beside=TRUE)
```

One plot that is helpful here.

#Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

No difference in values between x-axis and y-axis.

```{r}
pairs(x, col=mycols)
```

#Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?


# PCA to rescue 

Do PCR of this 17D food data. The main function in base R is called "prcom()". This function requires the transpose of our data in this case...

```{r}
pca <- prcomp(t(x))
summary(pca)
```

The prcom() function returns a list object. 

```{r}
plot(pca)
```

The PCA plot a.k.a a PCA score plot is a plot of PC1 vs PV2. Basically using the new PCA axis to view our data.

```{r}
attributes(pca)
```

We will focus on "pca$x" for this plot 

```{r}
pca$x
```

```{r}
plot(pca$x[,1], pca$x[,2], xlab= "PC1", ylab="PC2")
text(pca$x[,1], pca$x[,2], labels = colnames(x))
```
## PCA of a RNA-Seq

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

```{r}
pca <- prcomp(t(rna.data))
summary(pca)
```
Make our PCA score plot
```{r}
plot(pca$x[,1:2])
text(pca$x[,1:2], labels = colnames(rna.data))
```


