---
title: "Mini-project"
author: "Pham Vo"
date: "2/14/2022"
output:
  pdf_document: default
  html_document: default
---

```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
```

#Remove the first column as we will not use that for unsupervised analysis.
```{r}
wisc.data <- wisc.df[,-1]
```

# Create diagnosis vector for later 

```{r}
diagnosis <- wisc.df[,1]
```

> Q1. How many observations are in this dataset?

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df[,1]
ncol(wisc.data)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df[,1]
table(wisc.df$diagnosis)
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df[,1]
colvec <- colnames(wisc.data)
length(grep("_mean", colvec))
```

##Performing PCA

# Check column means and standard deviations

```{r}
colMeans(wisc.data)
apply(wisc.data,2,sd)
```

# Perform PCA on wisc.data by completing the following code

```{r}
wisc.data <- wisc.df[,-1]
wisc.data1 <- na.omit(wisc.data)
diagnosis <- wisc.df[,1]
wisc.data.scaled.2 <- data.frame(t(na.omit(t(wisc.data))))
wisc.pr <- prcomp(wisc.data.scaled.2, scale=TRUE)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.27%

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs

#Plot pca

```{r}
biplot(wisc.pr)
```

# Scatter plot observations by components 1 and 2

```{r}
plot(wisc.pr$x[,1:2], col = as.factor(diagnosis), xlab = "PC1", ylab = "PC2")
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(wisc.pr$x[,1:3], col = as.factor(diagnosis), xlab = "PC1", ylab = "PC3")
```

The first plot shows a better separation between the two groups.

> Plot using ggplot2 package

```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

# Variance explained

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var/sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")

```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

> Q9.  For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

-0.2608538 

```{r}
loading_scores <- wisc.pr$rotation[,1]
loading_scores["concave.points_mean"]
```


>Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

5 PCs.

# Hierarchical clustering

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

#Calculate the (Euclidean) distances between all pairs of observations in the new scaled dataset and assign the result to data.dist.

data.dist <- dist(data.scaled)

#Create a hierarchical clustering model using complete linkage. Manually specify the method argument to hclust() and assign the results to wisc.hclust.

wisc.hclust <- hclust(data.dist, method="complete")

#Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

plot(wisc.hclust)
abline(h=15, col="red", lty=2)
```

# Selecting number of clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)

#We can use the table() function to compare the cluster membership to the actual diagnoses.
table(wisc.hclust.clusters, diagnosis)
```
```{r}
tinytex::install_tinytex()
```






